STAT 672 Final Project:\\ Stochastic Gradient Descent

### 2018-04-17 paper/paper.tex
close to done

Consider a typical supervised classification problem. We have
feature matrix $\bm{X}_{n \times D}$ (corresponding to $n$ observations and $D$
features) and labels $\bm{Y}_{n \times 1}, Y_i \in \{-1, 1\}$. The goal is to
predict a particular observation's label $Y_i$ using that observation's features
$\bm{X}_i$. To emphasize the utility of SGD, suppose that both $n$ and $D$ are
very large.
We have a hypothesis class $\mathcal{F}$ consisting of
various functions $f_{\bm{w}}(\bm{X}_i) \in \mathcal{F}$ parametrized by weight
vector $\bm{w}$. We have convex loss function $L(Y_i, f_{\bm{w}}(\bm{X}_i))$ that
expresses the cost of misclassification. 
We will consider the optimal function $\hat{f}_{\bm{w}}(\bm{X}_i) \in F$ 
that which minimizes empirical risk over all observations: $\frac{1}{n}
\sum_{i=1}^n L(Y_i, f_{\bm{w}}(\bm{X}_i))$.
Denote $\hat{\bm{w}}$ the weight coefficients of this optimal function. We thus
have:

\begin{equation}
	\hat{\bm{w}} = \argmin_{\bm{w}}\frac{1}{n} \sum_{i=1}^n L(Y_i, f_{\bm{w}}(\bm{X}_i))
\end{equation}

@@

\begin{equation}
	\bm{w}^{(t+1)} = \bm{w}^{(t)} - \gamma \frac{1}{n}\sum_{i=1}^n
	\nabla_{\bm{w}} L(Y_i, f_{\bm{w}^{(t)}}(\bm{X}_i))
\end{equation}

@@

\begin{equation}
	\bm{w}^{(t+1)} = \bm{w}^{(t)} - \nabla_{\bm{w}} L(Y_i, f_{\bm{w}^{(t)}}(\bm{X}_i))
\end{equation}

Per previous analyses we know that this approach will converge to an arbitrarily
good answer (if certain assumptions are met); will converge to that arbitrary
degree of precision in less time than plain gradient descent; does not
necessarily require computing the Hessian; and can be parallelized across many
machines. These attractive qualities make SGD well-represented in almost all areas of
machine learning. In the following subsection, we take a closer look at one
particularly prominent applications, support vector machines.

@@

Originally proposed by \cite{boser1992training} and \cite{cortes1995support}, support vector machines (SVM)
are among the best supervised learning algorithms. Here we set up the SVM
problem and then show how SGD can be used to solve it.

\cite{shalev2011pegasos}

\subsection{Case Study}

\cite{dal2015calibrating}

\subsection{Real-World Applications}

>>>>

Consider a typical supervised classification problem. We have
feature matrix $\bm{X}_{n \times D}$ (corresponding to $n$ observations and $D$
features) and labels $\bm{Y}_{n \times 1}, Y_i \in \{-1, 1\}$. The goal is to
predict a particular observation's label $Y_i$ using that observation's features
$\bm{X}_i$. To emphasize the utility of SGD, suppose that both $n$ and $D$ are
very large.
We have a hypothesis class $\mathcal{F}$ consisting of
various functions $f(\bm{w}, \bm{X}_i) \in \mathcal{F}$ parametrized by weight
vector $\bm{w}$. We have convex loss function $L(Y_i, f(\bm{w}, \bm{X}_i))$ that
expresses the cost of misclassification. 
We will consider the optimal function
that which minimizes empirical risk over all observations: $\frac{1}{n}
\sum_{i=1}^n L(Y_i, f(\bm{w}, \bm{X}_i))$.
Denote $\hat{\bm{w}}$ the weight coefficients of this optimal function. We thus
have:

\begin{equation}
	\hat{\bm{w}} = \argmin_{\bm{w}}\frac{1}{n} \sum_{i=1}^n L(Y_i, f(\bm{w}, \bm{X}_i))
\end{equation}

@@

\begin{equation}
	\bm{w}^{(t+1)} = \bm{w}^{(t)} - \gamma \frac{1}{n}\sum_{i=1}^n
	\nabla_{\bm{w}} L(Y_i, f(\bm{w}^{(t)}, \bm{X}_i))
\end{equation}

@@

\begin{equation}
	\bm{w}^{(t+1)} = \bm{w}^{(t)} - \nabla_{\bm{w}} L(Y_i, f(\bm{w}^{(t)}, \bm{X}_i))
\end{equation}

Per previous analyses we know that this approach will converge to an arbitrarily
good answer (if certain assumptions are met); will converge to that arbitrary
degree of precision in less time than plain gradient descent; does not
necessarily require computing the Hessian; and can be parallelized across many
machines. These attractive qualities make SGD well-represented in almost all areas of
machine learning. In the following subsection, we take a closer look at one
particularly prominent application, support vector machines.

@@

Originally proposed by \cite{boser1992training} and \cite{cortes1995support}, support vector machines (SVM)
are among the best supervised learning algorithms. Here we set up the SVM
optimization
problem and then show how SGD can be used to solve it.

Suppose we have labels $\bm{Y}$ with $Y_i \in \{-1,1\}$ and features $\bm{X}_{n
\times D}$, corresponding to $n$ observations and $D$ features.
We want to construct a decision hyperplane 
$\langle \bm{w}, \bm{X} \rangle + w_0= 0$. $\bm{w}$ is
a weight vector that we apply to $\bm{X}$. If for a particular observation $i$
$\langle \bm{w}, \bm{X}_i \rangle + w_0 < 0$, we
classify $Y_i=-1$. Similarly, if $\langle \bm{w}, \bm{X}_i \rangle + w_0 > 0$, we
classify $Y_i = 1$. We introduce slack variables $\xi_i$ to allow some
violation of the margin, as well as a cost parameter $C$ to control them. The
primal soft-margin SVM problem thus is:

\begin{equation}
	\min_{\bm{w} \in \mathbb{R}^D, w_0 \in \mathbb{R}, \bm{\xi} \in \mathbb{R}^n} 
	\frac{1}{2} ||\bm{w}||^2 + C \frac{1}{n} \sum_{i=1}^n \xi_i
\end{equation}
$$
s.t. \quad Y_i(\langle \bm{w}, \bm{X}_i \rangle + w_0) \geq 1 - \xi_i 
$$
$$
\quad \xi_i \geq 0 
$$

This problem can be recharacterized through the lens of empirical risk
minimization. Suppose we obtain an optimal answer $\bm{w}^*, w_0^*, \bm{\xi}^*$. 
Then, it makes sense that:

$$
\xi_i^* = \max(0, 1 - Y_i(\langle \bm{w}, \bm{X}_i \rangle + w_0)) 
$$

We now can recharacterize the problem as:

$$
\min_{\bm{w}, w_0} C \frac{1}{n} \sum_{i=1}^n \left[ \max(0, 1 - Y_i(\langle \bm{w},
\bm{X}_i \rangle +
w_0)) + \frac{1}{2}||\bm{w}||^2 \right]
$$

This is equivalent to:
\begin{equation}
\min_{f \in \mathcal{F}} \frac{1}{n} \sum_{i=1}^n \left[ L(Y_i, \bm{w},
	\bm{X}_i) + \frac{\lambda}{2}
||\bm{w}||^2 \right]
\end{equation}

Where $\mathcal{F}$ is the linear hypothesis class, $L$ is the hinge loss

$$
L_{\mathrm{hinge}}(Y_i, \bm{w}, \bm{X}_i) = \max(0, 1 - Y_i
f(\bm{w}, \bm{X}_i))
$$
$$
f(\bm{w}, \bm{X}_i) = \bm{w}^T\bm{X}_i + w_0
$$

and
$\lambda=\frac{1}{2C}$. We use $\frac{\lambda}{2}$ for reasons of
convenience. This problem can be approached in different ways;
for example, it can be shown that the Lagrangian dual depends only on
``support vectors'' (observations violating the margin), leading to useful
data-reduction properties. Gradient descent is a
popular method to solve the primal formulation given here. First note
that the hinge loss function is not differentiable. We must use a sub-gradient:

\begin{equation}
	\frac{\partial L}{\partial \bm{w}} = 
	\begin{cases}
		-Y_i \bm{X}_i \quad \mathrm{if} \hspace{0.2em} Y_if(\bm{w}, \bm{X}_i) < 1\\
		0 \quad \mathrm{otherwise} \\
	\end{cases}
\end{equation}

The basic batch GD approach

$$
	\bm{w}^{(t+1)} = \bm{w}^{(t)} - \gamma \frac{1}{n} \sum_{i=1}^n
	\nabla_{\bm{w}} \left[ L(Y_i, \bm{w},
	\bm{X}_i) + \frac{\lambda}{2}
||\bm{w}||^2\right]
$$

$$
	= \bm{w}^{(t)} - \gamma \frac{1}{n} \sum_{i=1}^n
	 \left[ L\nabla_{\bm{w}} (Y_i, \bm{w},
	\bm{X}_i) + \lambda \bm{w} \right]
$$
thus becomes
\begin{equation}
	\bm{w}^{(t+1)} = 
	\begin{cases}
		\bm{w}^{(t)} - \gamma \frac{1}{n} \sum_{i=1}^n
		\left[
			\lambda \bm{w}^{(t)} - Y_i\bm{X}_i
		\right]
		\quad \mathrm{if} \hspace{0.2em} Y_if(\bm{w}^{(t)}, \bm{X}_i) < 1 \\

		\bm{w}^{(t)} - \gamma \frac{1}{n} \sum_{i=1}^n
		\lambda \bm{w}^{(t)} 
		\quad \textrm{otherwise} \\

	\end{cases}
\end{equation}

As previously discussed, batch GD is computationally expensive and
so for large $n$ we prefer SGD, selecting only a single randomly chosen
observation per iteration.

\begin{equation}
	\bm{w}^{(t+1)} = 
	\begin{cases}
		\bm{w}^{(t)} - \gamma 
		\left(
			\lambda \bm{w}^{(t)} - Y_i\bm{X}_i
		\right)
		\quad \mathrm{if} \hspace{0.2em} Y_if(\bm{w}^{(t)}, \bm{X}_i) < 1 \\

		\bm{w}^{(t)} - \gamma 
		\lambda \bm{w}^{(t)} 
		\quad \textrm{otherwise} \\

	\end{cases}
\end{equation}

There have been further efforts to improve SGD for the specific application of
SVM, most notably \cite{shalev2011pegasos}.

\subsection{Simulation Case Study}

Simulation provides a useful example of the computational advantages of SGD for
estimating coefficients in a predictive model. This section outlines such a
brief simulation, annotated with Python code.

Suppose we have feature matrix $\bm{X}_{n \times D}$ with $n=10^6$ and $D=1$
and outcome variable $\bm{Y}_i \sim N(\bm{X}_i,1)$.

We seek to estimate coefficients in a
regression equation of the form $y = \beta_0 + \beta_1 X$ by ordinary least
squares. We do so via two
competing methods: solving directly via $\hat{\bm{\beta}} =
(\bm{X}'\bm{X})^{-1}\bm{X}'\bm{Y}$ and iterative estimation via SGD.

We compare the estimated coefficients and run-time.
\subsection{Real-World Applications}
